Phylogenetics and the human microbiome.

### 2014-04-17 sbreview.tex
Capitulating to the microbiota-police.

The human microbiome is the collection of microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information from microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, but methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.

@@

\notforarxiv{
\vspace{4.5in}
\noindent (Keywords: human microbiome; microbial ecology; phylogenetic methods; 16S; metagenome)\\
\newpage
}

@@

The parameter regime and focus of microbiome research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods, because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.

@@

The human microbiome is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes is relatively rich.
It is also common to manipulate the human microbiome or animal models thereof via intervention studies and germ-free animals, which is easier to do than for many other microbial communities.

In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between microbial researchers and what I think of as the typical \emph{Systematic Biology} audience.
Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
I realize that there is substantial overlap--- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes--- but this terminology will be useful for concreteness.
There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
I will first briefly review the recent literature on the human microbiome, then describe novel ways in which human microbiome researchers have used trees.
I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics, indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.

\section{The human microbiome}
The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
The microbiome has remarkable metabolic potential, with an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiome, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
Our microbiomes have been historically transmitted vertically between mother and infant by seeding in the birth canal and breastfeeding;  immune dysfunction has been associated with the interruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
In this section I will briefly review what is known about the human microbiome and its effect on our health.

The human microbiome is an ecosystem.
It is dynamic in terms of taxonomic representation but apparently constant in terms of function \citep{hmp2012structure}.
There is a ``core'' microbiome which is shared between all humans \citep{turnbaugh2008core}.
The human microbiome is spatially organized, as can be seen on skin \citep{grice2009topographical}, with substantial variation in human body habitats across space and time \citep{costello2009bacterial}.
There is a substantial range of inter-individual versus intra-individual variation \citep{hmp2012structure}.

Our actions can shift the composition of our microbiome.
Changes in diet can very quickly shift its composition, but there is also a strong correlation between long-term diet and microbiome \citep{li2009human,wu2011linking}.
Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts for years \citep{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}.

The microbiome interacts on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
The gut microbiome in particular correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic dysfunction in rural Africa \citep{smith2013gut}.
Considerable attention has also been given to the interaction between the gut microbiome and obesity, although the story is not yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
A causal role for the microbiome leading to obesity is established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbiome transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiome leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
For example, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

Bacteria have been the primary focus of human microbiome research, and other domains have been investigated to a lesser extent.
Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea} and have a nonuniform distribution across the human skin \cite{Findley2013-rm}.
Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
We will focus on bacteria here.

In this paper we will primarily be describing the human microbiome from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiome will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.

@@

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a less than linear function.
\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiome; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

Nevertheless, the work that has been done concerning horizontal gene transfer in the human microbiome has revealed interesting results.
\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiome of Japanese such that individuals with this resulting microbiome are better able to digest the algae in their diet.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.
Horizontal transfer of genes is inferred in these studies by finding highly similar subsequences in otherwise less related organisms.

@@

Diversity preservation is of interest for microbiome researchers like it is for eukaryotic organisms, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms \citep{hartmann2006maximizing,pardi2007resource}.
Martin Blaser in particular has argued that changes in our microbiomes are leading to an increase in autoimmune disease and certain types of cancer \citep[reviewed in][]{cho2012human} and has made passionate appeals to preserve microbiome diversity \citep{blaser2011antibiotic}.
% Indeed, studies show that the modern lifestyle, including antibiotic use, has led to significant changes in microbiome.
Because a child's initial microbiome is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of microbiome extinction when the chain is interrupted.
Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiome development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiome from ancient feces.
How might phylogenetic methods be used in these preservation efforts?

Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiomes and their hosts.
\citet{ochman2010evolutionary} found identical tree topologies for primate and microbiome evolution.
For the microbiome, they used maximum parsimony such that each column was a microbe and each such entry took discrete states according to how much of that microbe was present.
Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiome change?
Would methods using these models do better than parsimony or commonly-applied phenetic methods applied to the distances described above?
Some studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
How can such forces be compared in this setting?

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiome?
At least for the next several years we can expect the research questions described above to continue to unfold.
Future research projects will continue to bring deeper sequencing on more samples.
The uBiome \footnote{\url{http://ubiome.com/}} and American Gut \footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
Comparative studies will continue to investigate what shapes and is shaped by the microbiome.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

Human microbiome research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
However, our microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiome analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has already contributed significantly to human microbiome research and will continue to do so.

>>>>

The human microbiome is the ensemble of genes in the microbes that live inside and on the surface of humans.
Because microbial sequencing information is now much easier to come by than phenotypic information, there has been an explosion of sequencing information from microbiome samples.
Much of the analytical work for these sequences involves phylogenetics, at least indirectly, but methodology has developed in a somewhat different direction than for other applications of phylogenetics.
In this paper I review the field and its methods from the perspective of a phylogeneticist, as well as describing current challenges for phylogenetics coming from this type of work.

@@

\notforarxiv{
\vspace{4.5in}
\noindent (Keywords: human microbiome; human microbiota; microbial ecology; phylogenetic methods; 16S; metagenome)\\
\newpage
}

@@

The parameter regime and focus of human-associated microbial research sits outside of the traditional setting for phylogenetics methods development and application; why should our community be interested in what microbial ecologists and medical researchers have done?
The answer is simple: this system is data- and question-rich.
It absolutely requires molecular methods, because microbes are now primarily identified by their molecular sequences, which is much more straightforward to do in high throughput than morphological or phenotypic characterization.
Indeed, microbial ecology has recently become for the most part the study of the relative abundances of various sequences derived from the environment, even if the framework for understanding between-microbe relationships includes metabolic information and other information not derived directly from sampled molecular sequences.

@@

The human-associated microbial assemblage is specifically interesting because questions of microbial genomics, translated into questions of function, have important consequences for human health.
Additionally, due to more than a century of hospital laboratory work, our knowledge about human-associated microbes is relatively rich.
This collection of microbes living inside and on top of us is called the \textit{human microbiota}, and the ensemble of genes in those microbes is called the \textit{human microbiome}, although usage of these terms varies \citep{Boon2013-mc}.

In this review I will describe phylogenetics-related research happening in microbial ecology and contrast approaches between microbial researchers and what I think of as the typical \emph{Systematic Biology} audience.
Despite an obvious oversimplification, I will use \textit{eukaryotic phylogenetics} to indicate what I think of as the mainstream of SB readership, and \textit{microbial phylogenetics} to denote the other.
I realize that there is substantial overlap--- for instance the microbial community is very interested in unicellular fungi, and additionally many in the SB community do work on microbes--- but this terminology will be useful for concreteness.
There is of course also substantial overlap in methodology, however as we will see there are significant differences in approach and the two areas have developed somewhat in parallel.
I will first briefly review the recent literature on the human microbiota, then describe novel ways in which human microbiome researchers have used trees.
I will finish with opportunities for the \textit{Systematic Biology} audience to contribute to this field.
I have made an explicit effort to make a neutral comparison between two directions rather than criticize the approximate methods common in microbial phylogenetics, indeed, microbial phylogenetics requires algorithms and ideas that work in parameter regimes an order of magnitude larger than typical for eukaryotic phylogenetics.

\section{The human microbiota}
The human microbiota is the collection of microbial organisms that live inside of and on the surface of humans.
These organisms are populous: it has been estimated that there are ten times as many bacteria associated with each individual than there are human cells of that individual.
The microbiota have remarkable metabolic potential, being an ensemble of genes estimated to be about 150 times larger than the human collection of genes \citep{qin2010human}.
Much of our metabolic interaction with the outside world is mediated by our microbiota, as it has important roles in immune system development, nutrition, and drug metabolism \citep{kau2011human,maurice2013xenobiotics}; our food and drug intake in turn impacts the diversity of microbes present.
Traditionally, our microbiota have been transmitted from mother to infant in the birth canal and by breastfeeding; immune dysfunction has been associated with the interruption of these practices \citep[reviewed in][]{funkhouser2013mom}.
In this section I will briefly review what is known about the human microbiota and its effect on our health.

The human microbiota form an ecosystem.
It is dynamic in terms of taxonomic representation but apparently constant in terms of function \citep{hmp2012structure}.
There is a ``core'' microbiota which is shared between all humans \citep{turnbaugh2008core}.
The human microbiota is spatially organized, as can be seen on skin \citep{grice2009topographical}, with substantial variation in human body habitats across space and time \citep{costello2009bacterial}.
There is a substantial range of inter-individual versus intra-individual variation \citep{hmp2012structure}.

Our actions can shift the composition of our microbiota.
Changes in diet can very quickly shift its composition, but there is also a strong correlation between long-term diet and microbiota \citep{li2009human,wu2011linking}.
Antibiotics fundamentally disturb microbial communities, resulting in an effect that lasts for years \citep{jernberg2007long,dethlefsen2008pervasive,jakobsson2010short,dethlefsen2011incomplete}.

The microbiota interact on many levels with host phenotype \citep[reviewed in][]{cho2012human}.
The gut microbiota in particular correlates with health of individuals from the elderly in industrialized nations \citep{claesson2012gut} to children with acute metabolic dysfunction in rural Africa \citep{smith2013gut}.
Considerable attention has also been given to the interaction between gut microbiota and obesity, although the story is not yet clear.
An intervention study has established human gut microbes associated with obesity \citep{ley2006microbial}.
A causal role for the microbiota leading to obesity has been established for mice: an obese phenotype can be transferred from mouse to mouse by gut microbial transplantation \citep{turnbaugh2006obesity}, the pregnant human gut microbiota leads to obesity in mice \citep{koren2012host}, and probiotics can lead to a lean phenotype and healthy eating behavior \citep{poutahidis2013microbial}.
However, these promising leads have not yet been confirmed causally or in population studies of humans \citep{zhao2013gut}.
For example, a study of obesity in the old-order Amish did not find any correlation between obesity and particular gut communities \citep{zupancic2012analysis}.

Bacteria have been the primary focus of human microbiota research, and other domains have been investigated to a lesser extent.
Changes in archaeal and fungal populations have been shown to covary with bacterial residents \citep{hoffmann2013archaea} and have a nonuniform distribution across the human skin \citep{Findley2013-rm}.
Viral populations have been observed to be highly dynamic and variable across individuals \citep{reyes2010viruses,minot2011human,minot2013rapid}.
We will focus on bacteria here.

In this paper we will primarily be describing the human microbiota from a community-level phylogenetic perspective rather than from the fine-scale perspective of immune-mediated interactions between host and microbe \citep[reviewed in][]{hooper2012interactions}.
Our understanding of the true effect of the microbiota will eventually come from such a molecular-level understanding, although until we can characterize all of the molecular interactions between microbes and the human body, a broad perspective will continue to be important.

@@

Although abundance weighted non-phylogenetic diversity measures such as \citet{simpson1949measurement} and \citet{shannon1948mathematical} are commonly used in human microbiome studies, abundance weighted phylogenetic diversity measures are not.
Abundance-weighted measures take a sum of branch lengths weighted by abundance, such that branches that connect abundant taxa get a higher weight than ones that do not (Fig.~\ref{fig:pd}).
Thus rare taxa and artifactual sequences are down-weighted compared to abundant taxa.
Such measures do exist \citep{rao1982diversity,barker2002phylogenetic,allen2009new,chao2010phylogenetic,vellend2011measuring}.
Abundance-weighted measures commonly weight edges in proportion to abundance, but one can also construct ``partially abundance weighted'' measures by weighting edges by abundances transformed by a less than linear function.
\citet{mccoy2013abundance} have recently shown that such partially abundance weighted diversity measures do a good job of distinguishing between dysbiotic and ``normal" states of the human microbiota; in particular that they do a better job than the commonly-used OTU-based measures.
\citet{nipperess2013mean} have also determined formulas for the expectation and variance of PD under random sub-sampling, which is often applied to enable comparison between samples of different sequencing depths and called ``rarefaction curves.''

@@

Nevertheless, the work that has been done concerning horizontal gene transfer in the human microbiome has revealed interesting results.
\citet{hehemann2010transfer} found that a seaweed gene has been transferred into a bacterium in the gut microbiota of Japanese such that individuals with this resulting microbiota are better able to digest the algae in their diet.
Following on this work, \citet{smillie2011ecology} found that the human microbiome is in fact a common location for gene transfer.
\citet{stecher2012gut} find that in a mouse model, horizontal transfer between pathogenic bacteria is blocked by commensal bacteria except for periods of gut inflammation.
Horizontal transfer of genes is inferred in these studies by finding highly similar subsequences in otherwise less related organisms.

@@

Diversity preservation is of interest for microbiota researchers like it is for eukaryotic organisms, but has not received the formalization and algorithmic treatment surrounding phylogenetic diversity for larger organisms \citep{hartmann2006maximizing,pardi2007resource}.
Martin Blaser in particular has argued that changes in our microbiota are leading to an increase in autoimmune disease and certain types of cancer \citep[reviewed in][]{cho2012human} and has made passionate appeals to preserve microbiota diversity \citep{blaser2011antibiotic}.
Because a child's initial microbiota is transmitted from the mother \citep[reviewed in][]{funkhouser2013mom}, there is a somewhat equivalent notion of microbiota extinction when the chain is interrupted.
Consequently, \citet{yatsunenko2012human} have explicitly contrasted microbiota development in urban, forest-dwelling, and rural populations, while \citet{tito2012insights} have endeavored to characterize the microbiota from ancient feces.
How might phylogenetic methods be used in these preservation efforts?

Perhaps in part because of the importance of mother-to-child transmission, there are indications of coevolution between microbiota and their hosts.
\citet{ochman2010evolutionary} found identical tree topologies for primate and microbiome evolution.
For the microbiota, they used maximum parsimony such that each column was a microbe and each such entry took discrete states according to how much of that microbe was present.
Although parsimony gave an interesting answer here, the presence of such coevolution raises the question of what sort of forward-time models are appropriate for microbiota change?
Would methods using these models do better than parsimony or commonly-applied phenetic methods applied to the distances described above?
Some studies \citep[e.g.][]{phillips2012microbiome,delsuc2013convergence} see a combination of historical and dietary influences.
How can such forces be compared in this setting?

@@

\section{Discussion}
What can we expect next at the intersection of phylogenetics and the human microbiota?
At least for the next several years we can expect the research questions described above to continue to unfold.
Future research projects will continue to bring deeper sequencing on more samples.
The uBiome \footnote{\url{http://ubiome.com/}} and American Gut \footnote{\url{http://americangut.org/}} projects promise to bring gut microbiome sequencing to the average citizen for a low price.
Comparative studies will continue to investigate what shapes and is shaped by the microbiota.
However, some of the initial excitement may have died down, as neither the Human Microbiome Project nor the MetaHIT project were extended.

@@

Human microbiota research has experienced a frenetic rate of expansion over the past decade, and sometimes the hype has outmatched the science.
However, our microbes are here to stay and so is research on them.
Thus we can look forward to the field of human microbiota analysis settling down to a comfortable and mature middle age as an interesting intersection between ecology and medicine.
Phylogenetics has already contributed significantly to research on the human microbiota and will continue to do so.
